---
title: "Individual Tumor Dynamics and Their Link to Survival: Two-Stage and Joint Modeling"
subtitle: "Portfolio project — Pharmacometrics-inspired modeling in oncology"
author: "Juliette Bouli-Mengue"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## 1. Motivation

In oncology drug development, individual tumor size trajectories (typically summarized
as Sum of Longest Diameters, SLD, per RECIST criteria) are often linked to survival
outcomes (Progression-Free Survival, PFS; Overall Survival, OS) to support dose
selection and efficacy predictions. This notebook demonstrates, on **fully simulated
data**, two complementary approaches used in pharmacometrics to link tumor dynamics to
survival: a classic "two-stage" approach (individual parameter estimation followed by a
Cox model), and a joint model (simultaneous estimation of the longitudinal trajectory
and survival).

No real patient data is used here. The goal is to demonstrate methodological
understanding before applying the same logic to real datasets.

### The Tumor Growth Inhibition (TGI) model

We use the biexponential TGI model (Wang et al., *Clin Pharmacol Ther*, 2009):

$$SLD(t) = SLD_0 \times (e^{-g \cdot t} + e^{d \cdot t} - 1)$$

- **g**: tumor regression rate (treatment effect)
- **d**: tumor regrowth rate (resistance)
- **SLD0**: baseline tumor size

```{r libraries}
library(minpack.lm)
library(survival)
library(survminer)
library(JM)
library(nlme)
library(dplyr)
library(ggplot2)

set.seed(42)
```

## 2. Simulating "true" individual patient parameters

We simulate 200 patients. Each has their own growth (`g`), regrowth (`d`), and baseline
tumor size (`SLD0`), drawn from log-normal distributions to reflect realistic
between-patient variability.

```{r simulate-patients}
n_patients <- 200

g_pop      <- 0.020
omega_g    <- 0.5
d_pop      <- 0.010
omega_d    <- 0.6
SLD0_pop   <- 60
omega_SLD0 <- 0.3

patients <- tibble(
  patient_id = 1:n_patients,
  g    = rlnorm(n_patients, log(g_pop), omega_g),
  d    = rlnorm(n_patients, log(d_pop), omega_d),
  SLD0 = rlnorm(n_patients, log(SLD0_pop), omega_SLD0)
)

head(patients)
```

## 3. Simulating survival outcomes

Survival is simulated using a proportional hazards structure: a higher `g` (better
response) protects the patient, while a higher `d` (faster regrowth) increases the
hazard.

```{r simulate-survival}
beta_g <- -15
beta_d <- 20
h0     <- 0.015

patients <- patients %>%
  mutate(
    g_c = g - mean(g),
    d_c = d - mean(d),
    linear_pred = beta_g * g_c + beta_d * d_c,
    hazard = h0 * exp(linear_pred),
    true_event_time = rexp(n(), rate = hazard)
  )

admin_censoring <- 104

patients <- patients %>%
  mutate(
    os_time  = pmin(true_event_time, admin_censoring),
    os_event = as.integer(true_event_time <= admin_censoring)
  )

summary(patients$os_time)
table(patients$os_event)
```

## 4. Simulating longitudinal tumor measurements

Tumor size is measured every 3 weeks over 72 weeks, with 8% multiplicative measurement
noise. *This observation design — scan frequency and measurement precision — was chosen
deliberately after an earlier, sparser design (6-week scans, 15% noise) failed to
recover a significant tumor-dynamics/survival association despite a strong true effect
by construction: a textbook case of attenuation by measurement error, where noisy
individual parameter estimates dilute a real underlying signal. The design below is the
one that successfully recovers the signal.*

```{r simulate-longitudinal}
assessment_times <- seq(0, 72, by = 3)
sigma_obs <- 0.08

longitudinal_data <- patients %>%
  rowwise() %>%
  do({
    pt <- .
    tibble(
      patient_id = pt$patient_id,
      time = assessment_times,
      SLD_true = pt$SLD0 * (exp(-pt$g * assessment_times) + exp(pt$d * assessment_times) - 1)
    )
  }) %>%
  ungroup() %>%
  mutate(SLD_obs = SLD_true * exp(rnorm(n(), 0, sigma_obs))) %>%
  filter(SLD_obs > 0)

ggplot(longitudinal_data %>% filter(patient_id <= 12),
       aes(x = time, y = SLD_obs, group = patient_id, color = factor(patient_id))) +
  geom_line() + geom_point() +
  labs(title = "Simulated tumor size trajectories", x = "Weeks", y = "SLD (mm)", color = "Patient") +
  theme_minimal()
```

## 5. Two-stage approach: re-estimating g and d from noisy scans

We act as if the true `g` and `d` were unknown, and re-estimate them for each patient
from their noisy longitudinal measurements only.

```{r two-stage-fit}
fit_one_patient <- function(df) {
  tryCatch({
    fit <- nlsLM(
      SLD_obs ~ SLD0_est * (exp(-g_est * time) + exp(d_est * time) - 1),
      data = df,
      start = list(SLD0_est = df$SLD_obs[df$time == 0][1], g_est = 0.02, d_est = 0.01),
      lower = c(SLD0_est = 1, g_est = 0.0001, d_est = 0.0001),
      upper = c(SLD0_est = 500, g_est = 1, d_est = 1),
      control = nls.lm.control(maxiter = 200)
    )
    coefs <- coef(fit)
    tibble(g_hat = coefs["g_est"], d_hat = coefs["d_est"], convergence = TRUE)
  }, error = function(e) {
    tibble(g_hat = NA_real_, d_hat = NA_real_, convergence = FALSE)
  })
}

individual_fits <- longitudinal_data %>%
  group_by(patient_id) %>%
  filter(n() >= 4) %>%
  group_modify(~ fit_one_patient(.x)) %>%
  ungroup()

cat("Convergence rate:", mean(individual_fits$convergence, na.rm = TRUE) * 100, "%\n")

analysis_data <- patients %>%
  inner_join(individual_fits, by = "patient_id") %>%
  filter(convergence, !is.na(g_hat), !is.na(d_hat))

cox_model <- coxph(Surv(os_time, os_event) ~ g_hat + d_hat, data = analysis_data)
summary(cox_model)

cor(analysis_data$g, analysis_data$g_hat)
cor(analysis_data$d, analysis_data$d_hat)
```

```{r two-stage-plot}
ggplot(analysis_data, aes(x = g, y = g_hat)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Estimated vs true g", x = "True g (hidden)", y = "Estimated g (g_hat)") +
  theme_minimal()
```

**Interpretation.** `g_hat` and `d_hat` recover the true parameters well (correlations
of 0.83 and 0.99 respectively), and the Cox model detects a significant association
between `g_hat` and survival (p = 0.007), consistent with the true simulated effect
(`beta_g = -15`).

The two-stage approach, however, has a structural limitation: it treats parameter
estimation and survival modeling as two independent steps, ignoring the uncertainty of
`g_hat`/`d_hat` when they are reused as fixed covariates. This motivates the joint
modeling approach developed below.

```{r save-for-later-notebooks}
saveRDS(patients, "patients_final.rds")
saveRDS(longitudinal_data, "longitudinal_data_final.rds")
```

## 6. Joint model: simultaneous estimation

A joint model estimates the longitudinal trajectory and survival simultaneously,
sharing information through common random effects, with estimation uncertainty
properly propagated into the survival model. Standard R joint modeling packages (`JM`)
require a linear (not biexponential) longitudinal submodel; we therefore model
`log(SLD)` as a linear function of time — a deliberate, documented simplification.

```{r prepare-joint-data}
joint_data <- longitudinal_data %>%
  filter(SLD_obs > 0) %>%
  mutate(log_SLD = log(SLD_obs)) %>%
  inner_join(patients %>% dplyr::select(patient_id, os_time, os_event), by = "patient_id") %>%
  mutate(patient_id = factor(patient_id))

surv_data <- patients %>%
  mutate(patient_id = factor(patient_id)) %>%
  as.data.frame()
```

```{r fit-joint-model}
fit_lme <- lme(log_SLD ~ time, random = ~ time | patient_id, data = joint_data)

fit_cox <- coxph(Surv(os_time, os_event) ~ 1, data = surv_data, x = TRUE)

fit_joint <- jointModel(fit_lme, fit_cox, timeVar = "time", method = "piecewise-PH-aGH")

summary(fit_joint)
```

**Interpretation.** The association parameter ("Assoct") quantifies the link between
the current value of the longitudinal trajectory and the hazard. Here, `Assoct ≈ 0.235`
(p ≈ 0.056): a higher current `log(SLD)` is associated with increased hazard, in the
biologically expected direction, though only borderline significant.

## 7. Two-stage vs. joint model: an honest comparison

| Approach | Signal detected | Model complexity | Trade-off |
|---|---|---|---|
| Two-stage | Significant (p = 0.007) on `g_hat` | Rich (two biological rate parameters, g and d) | Ignores estimation uncertainty of g_hat/d_hat |
| Joint model | Borderline (p = 0.056) | Simplified (single linear trend) | Statistically more rigorous, but coarser biological representation |

The joint model is not "worse" — it is more conservative because it properly accounts
for trajectory uncertainty, using a deliberately simplified tumor dynamics
representation. This is a genuine methodological trade-off: pharmacometricians must
balance model realism against statistical tractability and interpretability.

## 8. Limitations

- The two-stage approach ignores estimation uncertainty when parameters are reused as
  fixed covariates; the joint model addresses this but uses a simplified linear tumor
  trajectory instead of the full biexponential model.
- The association structure used in the joint model is the "current value"
  parameterization; other parameterizations (current slope, cumulative effects) could
  be explored and compared.
- Validation on a real dataset (e.g., a public NSCLC trial from Project Data Sphere)
  would be the natural next step, once methodological understanding is demonstrated on
  simulated data with known ground truth.
