---
title: "Joint Modeling of Tumor Dynamics and Survival"
subtitle: "Portfolio project — Pharmacometrics-inspired modeling in oncology (part 2)"
author: "Juliette Bouli-Mengue"
output: html_document
---

```{r setup2, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## 1. Why a joint model?

In the companion notebook (`01_simulation_and_two_stage_model.Rmd`), tumor dynamics
parameters (`g`, `d`) were first estimated independently for each patient, then reused as
fixed covariates in a Cox model. This "two-stage" approach ignores the uncertainty
attached to the individual parameter estimates, which can attenuate the estimated
association with survival (Rizopoulos, *Joint Models for Longitudinal and Time-to-Event
Data*, 2012).

A **joint model** estimates the longitudinal trajectory and the survival process
**simultaneously**, sharing information between the two through common random effects.
The current value (or trend) of the longitudinal trajectory becomes a time-dependent
predictor of the hazard, with its estimation uncertainty properly propagated into the
survival model.

### A necessary simplification

Standard joint modeling packages in R (`JM`, `joineR`, `JMbayes2`) are built around
linear or simple polynomial longitudinal trajectories, not around complex nonlinear
models such as the full biexponential TGI equation used in notebook 1. For this reason,
we simplify the tumor trajectory to a **linear trend on the log scale**:
`log(SLD)` as a linear function of time, with a patient-specific slope and intercept.

This is a deliberate, documented simplification: rather than modeling tumor regression
and regrowth as two separate biological rates, we summarize the overall trend of each
patient's tumor burden and link that trend directly to survival — enough to demonstrate
the joint modeling principle without requiring a highly specialized nonlinear joint model
implementation.

```{r libraries2}
library(JM)
library(nlme)
library(survival)
library(dplyr)
library(ggplot2)

set.seed(42)
```

*(This notebook assumes `longitudinal_data` and `patients` were generated using the
improved design from notebook 1: 3-week scans, 8% measurement noise, over 72 weeks.)*

## 2. Preparing the data

The `JM` package requires two datasets:

- a **long-format** dataset for the longitudinal submodel (one row per measurement),
  including survival information on each row
- a **one-row-per-patient** dataset for the survival submodel


```{r load-data}
patients <- readRDS("patients_final.rds")
longitudinal_data <- readRDS("longitudinal_data_final.rds")
```

```{r prepare-data}
joint_data <- longitudinal_data %>%
  filter(SLD_obs > 0) %>%
  mutate(log_SLD = log(SLD_obs)) %>%
  inner_join(patients %>% dplyr::select(patient_id, os_time, os_event), by = "patient_id") %>%
  mutate(patient_id = factor(patient_id))

head(joint_data)

surv_data <- patients %>%
  mutate(patient_id = factor(patient_id)) %>%
  as.data.frame()
```

## 3. Fitting the longitudinal submodel

We model `log(SLD)` as a linear function of time, with a random intercept and random
slope per patient — the "mixed" part of the model captures both the population-average
trend and each patient's individual deviation from it.

```{r fit-lme}
fit_lme <- lme(
  log_SLD ~ time,
  random = ~ time | patient_id,
  data = joint_data
)

summary(fit_lme)
```

**Interpretation.** The fixed effect for `time` (~0.0059, p < 0.001) indicates that, on
average across the simulated population, `log(SLD)` slightly *increases* over the 72-week
follow-up. This is not a sign of treatment failure: recall the underlying biexponential
shape (initial shrinkage from `g`, followed by regrowth from `d`) — over this full
observation window, the regrowth phase numerically dominates the average trend. The
random effects show substantial between-patient variability in both baseline tumor size
and individual slope — this individual slope variability is what we will now connect to
survival.

## 4. Fitting the survival submodel

```{r fit-cox-null}
fit_cox <- coxph(
  Surv(os_time, os_event) ~ 1,
  data = surv_data,
  x = TRUE
)
```

The Cox submodel is intentionally fit without covariates (`~ 1`): the link between
tumor trajectory and hazard is introduced automatically by the joint model in the next
step, not manually here.

## 5. Assembling the joint model

```{r fit-joint}
fit_joint <- jointModel(
  fit_lme,
  fit_cox,
  timeVar = "time",
  method = "piecewise-PH-aGH"
)

summary(fit_joint)
```

## 6. Interpreting the association parameter

The key output of a joint model is the **association parameter** ("Assoct"), which
quantifies the strength of the link between the current value of the longitudinal
trajectory and the hazard of the event — estimated with the corresponding uncertainty
of the trajectory itself, rather than a single fixed point estimate as in the two-stage
approach.

In our fit, `Assoct ≈ 0.235` (p ≈ 0.056): a higher current `log(SLD)` (i.e., a larger
tumor burden at a given time) is associated with an increased hazard of death, in the
biologically expected direction, though only borderline significant at the conventional
0.05 threshold.

## 7. Two-stage vs. joint model: an honest comparison

| Approach | Signal detected | Model complexity | Trade-off |
|---|---|---|---|
| Two-stage (improved design) | Significant (p = 0.007) on `g_hat` | Rich (two biological rate parameters, g and d) | Ignores estimation uncertainty of g_hat/d_hat |
| Joint model | Borderline (p = 0.056) on the association parameter | Simplified (single linear trend) | Statistically more rigorous, but coarser biological representation |

The joint model is not "worse" here — it is more conservative because it properly
accounts for the uncertainty in the estimated trajectory, and it uses a deliberately
simplified representation of tumor dynamics (a single slope rather than separate growth
and regrowth rates). This is a genuine methodological trade-off, not a failure of either
approach: it illustrates why, in practice, pharmacometricians must balance model realism
against statistical tractability and interpretability.

## 8. Limitations and next steps

- The longitudinal submodel here uses a linear trend on the log scale; a natural
  extension would be a nonlinear joint model directly incorporating the biexponential
  TGI structure (growth and regrowth rates), at the cost of substantially higher
  implementation complexity.
- The association structure used here is the "current value" parameterization; other
  parameterizations (current slope, cumulative effects) could be explored and compared.
- Validation on a real dataset (e.g., a public NSCLC trial from Project Data Sphere)
  would be the natural next step, once methodological understanding is demonstrated on
  simulated data with known ground truth.
