---
title: "Multi-Lesion RECIST Progression and PFS/OS Modeling"
subtitle: "Pharmacometrics-inspired modeling in oncology (part 3)"
author: "Juliette Bouli-Mengue"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## 1. Motivation

Notebooks 1 and 2 modeled a single aggregate tumor size (SLD) per patient. In reality, RECIST criteria are based on the **sum** of several individually measured target lesions (up to 5), each of which may respond differently to treatment even within the same patient. This notebook extends the framework by:

  - simulating multiple individual lesions per patient,
  - deriving total SLD as their sum,
  - detecting disease progression using RECIST 1.1 rules (rather than treating   progression as an abstract hazard, as in previous notebooks),
  - constructing Progression-Free Survival (PFS) as a composite time-to-event endpoint (the earliest of RECIST progression or death),
  - comparing PFS and OS on the same simulated patients.

```{r libraries}
library(dplyr)
library(ggplot2)
library(survival)
library(survminer)

set.seed(42)
```

## 2. Loading the simulated population

We reuse the simulated oncology population generated in notebook 1: each patient has
individual tumor growth parameters (`g`, `d`) representing their overall disease
trajectory. These patient-level parameters will serve as the basis for simulating
several individual lesions per patient in Section 3.

```{r load-data}
# Assumes patients_final.rds is saved in the same folder as this notebook
# (produced by notebook 1), a relative path, so the project stays portable across machines.
patients <- readRDS("patients_final.rds")

head(patients)
```

## 3. Simulating multiple lesions per patient

RECIST-based tumor burden is the **sum** of several individually measured target lesions (up to 5 in clinical practice), each of which may respond differently to
treatment even within the same patient. 
Rather than continuing to treat SLD as a single aggregate value per patient, we simulate 5 to 10 individual lesions per patient, each with its own growth (`g_lesion`) and regrowth (`d_lesion`) rate, and its own baseline size (`SLD0_lesion`). All derived from the patient-level parameters (`g`, `d`, `SLD0`) from notebook 1, with lesion-specific variation applied on top.

```{r create-lesions}
n_lesions_min <- 5
n_lesions_max <- 10

lesions <- patients %>%
  rowwise() %>%
  do({
    pt <- .
    n_lesions <- sample(n_lesions_min:n_lesions_max, 1)

    tibble(
      patient_id = pt$patient_id,
      lesion_id = 1:n_lesions,
      g_lesion = pt$g * runif(n_lesions, 0.7, 1.3),
      d_lesion = pt$d * runif(n_lesions, 0.7, 1.3),
      SLD0_lesion = (pt$SLD0 / n_lesions) * runif(n_lesions, 0.7, 1.3)
    )
  }) %>%
  ungroup()

head(lesions, 15)
cat("Number of lesions per patient,  summary:\n")
summary(table(lesions$patient_id))

lesions %>%
  group_by(patient_id) %>%
  summarise(n_lesions = n()) %>%
  summarise(min_lesions = min(n_lesions), max_lesions = max(n_lesions), mean_lesions = mean(n_lesions))
```

## 4. Simulating individual lesion trajectories and total tumor burden

Each lesion follows the same biexponential TGI equation used in notebook 1, but with its own lesion-specific parameters instead of the patient-level ones. 
We simulate scans every 6 weeks over 104 weeks, matching the follow-up duration used for the survival outcomes in notebook 1, then sum all lesions belonging to the same patient at each time point to obtain total SLD, exactly as RECIST defines it.

```{r create-trajectories}
assessment_times <- seq(0, 104, by = 6)

lesion_trajectories <- lesions %>%
  rowwise() %>%
  do({
    ls <- .
    tibble(
      patient_id = ls$patient_id,
      lesion_id = ls$lesion_id,
      time = assessment_times,
      diameter = ls$SLD0_lesion * (exp(-ls$g_lesion * assessment_times) + exp(ls$d_lesion * assessment_times) - 1)
    )
  }) %>%
  ungroup()

total_sld <- lesion_trajectories %>%
  group_by(patient_id, time) %>%
  summarise(SLD_total = sum(diameter), .groups = "drop")

head(total_sld, 12)

```
## 5. Detecting RECIST progression

At each scan, we compare the current total SLD to the **nadir**, the lowest SLD value observed for that patient up to and including that time point. 
Progression is declared the first time both RECIST conditions are met simultaneously: at least a 20% increase
from nadir, and at least a 5mm absolute increase from nadir.

```{r detect-progression}
recist_data <- total_sld %>%
  group_by(patient_id) %>%
  arrange(time, .by_group = TRUE) %>%
  mutate(
    nadir = cummin(SLD_total),
    relative_increase = (SLD_total - nadir) / nadir,
    absolute_increase = SLD_total - nadir,
    is_progression = relative_increase >= 0.20 & absolute_increase >= 5
  ) %>%
  ungroup()

head(recist_data, 12)
recist_data %>% filter(patient_id == 4) %>% select(time, SLD_total, nadir, relative_increase, absolute_increase, is_progression)
```

## 6. Extracting time to RECIST progression

For each patient, we identify the first time point at which RECIST progression is detected. Patients who never meet the progression criteria within the follow-up window
are treated as censored for progression at their last observed time point.

```{r extract-progression-times}
progression_times <- recist_data %>%
  group_by(patient_id) %>%
  summarise(
    progression_time = if (any(is_progression)) min(time[is_progression]) else NA_real_,
    progression_event = as.integer(any(is_progression)),
    last_observed_time = max(time),
    .groups = "drop"
  )

head(progression_times, 10)
cat("Number of patients with RECIST progression detected:", sum(progression_times$progression_event), "out of", nrow(progression_times), "\n")
progression_times %>% filter(progression_event == 0)
```

## 7. Constructing PFS as a composite endpoint

Progression-Free Survival (PFS) is a composite time-to-event endpoint: a patient experiences a PFS event at the earliest of RECIST progression or death, whichever occurs first. We combine the RECIST progression times from Section 6 with the OS data from notebook 1 to construct PFS.

```{r construct-pfs}
pfs_data <- patients %>%
  select(patient_id, os_time, os_event) %>%
  left_join(progression_times, by = "patient_id") %>%
  mutate(
    progression_time_filled = if_else(progression_event == 1, progression_time, last_observed_time),
    pfs_time = pmin(progression_time_filled, os_time),
    pfs_event = as.integer(progression_event == 1 | os_event == 1)
  )

head(pfs_data, 10)
```

## 8. Comparing PFS and OS

As expected clinically, PFS events typically occur earlier than OS events, since disease progression is usually detected before death. We visualize both endpoints on the same population using Kaplan-Meier curves.

```{r compare-pfs-os}
comparison_data <- bind_rows(
  pfs_data %>% transmute(patient_id, time = pfs_time, event = pfs_event, endpoint = "PFS"),
  pfs_data %>% transmute(patient_id, time = os_time, event = os_event, endpoint = "OS")
)

km_comparison <- survfit(Surv(time, event) ~ endpoint, data = comparison_data)

ggsurvplot(km_comparison, data = comparison_data, risk.table = TRUE, pval = FALSE,
           xlab = "Time (weeks)", ylab = "Survival probability",
           title = "PFS vs. OS on the same simulated population",
           legend.title = "Endpoint")

comparison_data %>%
  group_by(endpoint) %>%
  summarise(median_time = median(time), event_rate = mean(event))
```

## 9. Conclusion and limitations

This notebook extends the tumor dynamics framework from notebooks 1 and 2 to a
multi-lesion structure, closer to how RECIST is actually applied in clinical trials:
total tumor burden as the sum of individually simulated lesions, progression detected
via explicit RECIST 1.1 rules, and PFS constructed as a genuine composite endpoint
rather than an independently simulated hazard.

**Limitations, stated explicitly:**

- Lesions are simulated as varying independently around the patient-level `g` and `d`
  from notebook 1, rather than being generated from a shared mechanistic process; this
  is a simplification of true inter-lesion heterogeneity (e.g., differential drug
  penetration or clonal resistance across metastatic sites).
- Scans continue at fixed 6-week intervals throughout the full follow-up window
  regardless of death, which is not clinically realistic (a deceased patient has no
  further scans); this does not bias `pfs_time` itself, since `pmin()` always resolves
  to the earlier of progression or death, but it is a modeling simplification worth
  noting.
- New lesion appearance — one of the three RECIST progression criteria alongside target
  lesion growth — is not modeled here; only growth of existing target lesions is
  considered.

