---
title: "Does the ADC Improve PFS and OS? A Multi-Lesion RECIST Comparison"
subtitle: "Pharmacometrics-inspired modeling in oncology (part 4)"
author: "Juliette Bouli-Mengue"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## 1. Motivation

Notebook 3 introduced multi-lesion RECIST progression and PFS as a composite endpoint,
using the biexponential growth/regrowth model from notebook 1 (no drug mechanism).
Notebook 2 introduced an explicit ADC pharmacokinetic/pharmacodynamic mechanism, but
worked with a single aggregate tumor size and a simplified landmark survival analysis.

This notebook combines both: individual lesions (as in notebook 3) are driven by the
ADC PK/PD mechanism (as in notebook 2), allowing us to compare PFS and OS between a
treated (ADC) arm and an untreated counterfactual arm, using the same RECIST-based
progression detection built in notebook 3.

**Two compatibility choices, made explicitly:**

- **Time scale**: this notebook works in **days** (as in notebook 2), extended to 728
  days (~104 weeks) to allow enough follow-up time for progression and death events, as
  in notebook 3.
- **Growth model**: lesions here follow the simpler growth-minus-drug-effect structure
  from notebook 2 (`dSLD/dt = g*SLD - Effect(t)*SLD`), not the biexponential model from
  notebooks 1/3, since the drug-effect counterfactual logic was already validated on
  this structure in notebook 2.
  
```{r libraries}
library(dplyr)
library(ggplot2)
library(survival)
library(survminer)

set.seed(42)
```  

## 2. Loading the simulated population

We reuse the simulated oncology population generated in notebook 1: each patient has
individual tumor growth parameters (`g`, `d`) representing their overall disease
trajectory. These patient-level parameters will serve as the basis for simulating
several individual lesions per patient in Section 3.

```{r load-data}
# Assumes patients_final.rds is saved in the same folder as this notebook
# (produced by notebook 1), a relative path, so the project stays portable across machines.
patients <- readRDS("patients_final.rds")

head(patients)
```

## 3. Simulating multiple lesions per patient

As in notebook 3, each patient has 5 to 10 individual lesions. Unlike notebook 3,lesions here do not carry a `d` (intrinsic regrowth) parameter, tumor regrowth in this notebook emerges solely from the ADC effect fading over time (Section 5), following the mechanism validated in notebook 2.

```{r create-lesions}
n_lesions_min <- 5
n_lesions_max <- 10

lesions <- patients %>%
  rowwise() %>%
  do({
    pt <- .
    n_lesions <- sample(n_lesions_min:n_lesions_max, 1)

    tibble(
      patient_id = pt$patient_id,
      lesion_id = 1:n_lesions,
      g_lesion = pt$g * runif(n_lesions, 0.7, 1.3),
      SLD0_lesion = (pt$SLD0 / n_lesions) * runif(n_lesions, 0.7, 1.3)
    )
  }) %>%
  ungroup()

head(lesions, 10)
```

## 4. Applying PK/PD: repeated dosing, one drug exposure profile per patient

Real ADC regimens involve repeated administration (typically every 21 days), not a single dose. We model plasma concentration using the superposition principle: total concentration at any time is the sum of contributions from all doses already administered, each decaying independently from its own injection time. This produces the characteristic sawtooth accumulation pattern of repeated dosing, and sustains drug exposure, and therefore pharmacodynamic effect, over a much longer period than a single injection would. Because repeated dosing itself now sustains the effect, we revert `Emax` and `kel` to the same realistic values used in notebook 2, rather than artificially inflating them.

```{r pk-pd-setup}
lesions <- lesions %>%
  mutate(g_daily = g_lesion / 7)

dose <- 5     # mg/kg
V    <- 3     # L/kg
EC50 <- 1
Emax <- 0.003

patients <- patients %>%
  mutate(kel_i = rlnorm(n(), meanlog = log(0.05), sdlog = 0.20))

time_grid <- seq(0, 728, by = 7)  # days, weekly scans over ~104 weeks
dosing_interval <- 21  # days, typical q3w ADC regimen
dosing_times <- seq(0, 728, by = dosing_interval)

compute_concentration <- function(t_vec, kel, dosing_times, dose, V) {
  sapply(t_vec, function(t) {
    active_doses <- dosing_times[dosing_times <= t]
    if (length(active_doses) == 0) return(0)
    sum((dose / V) * exp(-kel * (t - active_doses)))
  })
}

patient_pkpd <- patients %>%
  rowwise() %>%
  do({
    pt <- .
    conc <- compute_concentration(time_grid, pt$kel_i, dosing_times, dose, V)
    inhibition <- Emax * conc / (EC50 + conc)
    tibble(patient_id = pt$patient_id, time = time_grid, concentration = conc, inhibition = inhibition)
  }) %>%
  ungroup()

ggplot(patient_pkpd %>% filter(patient_id == 1), aes(x = time, y = concentration)) +
  geom_line() +
  labs(title = "ADC concentration with repeated q3w dosing (patient 1)",
       x = "Time (days)", y = "ADC concentration") +
  theme_minimal()


head(patients %>% select(patient_id, kel_i))
```
## 5. Simulating each lesion under repeated-dosing treatment and the no-drug counterfactual

Each lesion now uses its patient's repeated-dosing inhibition profile (Section 4, computed once per patient) rather than a single-dose profile computed per lesion. As before, we simulate two trajectories per lesion: one under ADC treatment, and one no-drug counterfactual.

```{r simulate-lesion-trajectories}
lesion_trajectories <- data.frame()

for (i in 1:nrow(lesions)) {
  pid    <- lesions$patient_id[i]
  g_i    <- lesions$g_daily[i]
  SLD0_i <- lesions$SLD0_lesion[i]

  inhibition <- patient_pkpd$inhibition[patient_pkpd$patient_id == pid]

  SLD_adc     <- numeric(length(time_grid)); SLD_adc[1] <- SLD0_i
  SLD_no_drug <- numeric(length(time_grid)); SLD_no_drug[1] <- SLD0_i

  for (t in 2:length(time_grid)) {
    SLD_adc[t]     <- SLD_adc[t - 1] + (g_i * SLD_adc[t - 1] - inhibition[t - 1] * SLD_adc[t - 1])
    SLD_no_drug[t] <- SLD_no_drug[t - 1] + (g_i * SLD_no_drug[t - 1])
  }

  lesion_trajectories <- bind_rows(
    lesion_trajectories,
    data.frame(patient_id = pid, lesion_id = lesions$lesion_id[i], time = time_grid,
               SLD_adc = SLD_adc, SLD_no_drug = SLD_no_drug)
  )
}

head(lesion_trajectories, 10)
```
## 6. Total tumor burden per arm

We sum all lesions belonging to the same patient, separately for each arm (ADC and no-drug counterfactual), to obtain total SLD over time — the RECIST-based tumor burden metric used for progression detection in Section 7.

```{r total-sld-by-arm}
total_sld_by_arm <- lesion_trajectories %>%
  group_by(patient_id, time) %>%
  summarise(
    SLD_total_adc = sum(SLD_adc),
    SLD_total_no_drug = sum(SLD_no_drug),
    .groups = "drop"
  )

head(total_sld_by_arm, 10)
```
## 7. Detecting RECIST progression on both arms

We apply the same RECIST 1.1 progression rule from notebook 3 (≥20% increase from nadir AND ≥5mm absolute increase) independently to each arm, since the two arms represent two different counterfactual disease courses for the same patient.

```{r detect-progression-both-arms}
detect_progression <- function(data, sld_column) {
  data %>%
    group_by(patient_id) %>%
    arrange(time, .by_group = TRUE) %>%
    mutate(
      nadir = cummin(.data[[sld_column]]),
      relative_increase = (.data[[sld_column]] - nadir) / nadir,
      absolute_increase = .data[[sld_column]] - nadir,
      is_progression = relative_increase >= 0.20 & absolute_increase >= 5
    ) %>%
    ungroup()
}

recist_adc <- detect_progression(total_sld_by_arm, "SLD_total_adc")
recist_no_drug <- detect_progression(total_sld_by_arm, "SLD_total_no_drug")

head(recist_adc %>% select(patient_id, time, SLD_total_adc, nadir, is_progression), 10)
```
## 8. Constructing PFS and OS for both arms

For each arm, we extract the first time of RECIST progression (as in notebook 3, Section 6). For OS, we reuse the landmark-based counterfactual hazard approach validated in notebook 2: both arms share the same hazard structure, differing only in the tumor burden (at the last observed time point) that feeds into it, ensuring a fair comparison, not two unrelated survival models.

```{r construct-pfs-both-arms}
extract_progression_times <- function(recist_df) {
  recist_df %>%
    group_by(patient_id) %>%
    summarise(
      progression_time = if (any(is_progression)) min(time[is_progression]) else NA_real_,
      progression_event = as.integer(any(is_progression)),
      last_observed_time = max(time),
      .groups = "drop"
    )
}

prog_adc <- extract_progression_times(recist_adc)
prog_no_drug <- extract_progression_times(recist_no_drug)

# Final tumor burden per arm (last observed time point), used as the OS prognostic biomarker
final_sld <- total_sld_by_arm %>%
  filter(time == max(time)) %>%
  select(patient_id, SLD_total_adc, SLD_total_no_drug)

baseline_hazard <- 0.002
beta_SLD <- 0.01
pooled_mean_sld <- mean(c(final_sld$SLD_total_adc, final_sld$SLD_total_no_drug))

os_data <- bind_rows(
  final_sld %>% transmute(patient_id, SLD_used = SLD_total_adc, arm = "ADC"),
  final_sld %>% transmute(patient_id, SLD_used = SLD_total_no_drug, arm = "No ADC")
) %>%
  mutate(
    hazard = baseline_hazard * exp(beta_SLD * (SLD_used - pooled_mean_sld)),
    event_time = rexp(n(), rate = hazard),
    admin_censoring = 728,
    os_time = pmin(event_time, admin_censoring),
    os_event = as.integer(event_time <= admin_censoring)
  )

pfs_data_adc <- prog_adc %>%
  left_join(os_data %>% filter(arm == "ADC") %>% select(patient_id, os_time, os_event), by = "patient_id") %>%
  mutate(
    progression_time_filled = if_else(progression_event == 1, progression_time, last_observed_time),
    pfs_time = pmin(progression_time_filled, os_time),
    pfs_event = as.integer(progression_event == 1 | os_event == 1),
    arm = "ADC"
  )

pfs_data_no_drug <- prog_no_drug %>%
  left_join(os_data %>% filter(arm == "No ADC") %>% select(patient_id, os_time, os_event), by = "patient_id") %>%
  mutate(
    progression_time_filled = if_else(progression_event == 1, progression_time, last_observed_time),
    pfs_time = pmin(progression_time_filled, os_time),
    pfs_event = as.integer(progression_event == 1 | os_event == 1),
    arm = "No ADC"
  )

pfs_data_combined <- bind_rows(pfs_data_adc, pfs_data_no_drug)

head(pfs_data_combined, 10)
pfs_data_combined %>% filter(patient_id %in% c(1, 2)) %>% select(patient_id, arm, progression_time, os_time, pfs_time, pfs_event)
```
## 9. Comparing PFS and OS between ADC and No ADC arms

We now visualize whether the ADC treatment improves both PFS and OS relative to the no-drug counterfactual, using Kaplan-Meier curves for each endpoint.

```{r compare-arms-km}
km_pfs <- survfit(Surv(pfs_time, pfs_event) ~ arm, data = pfs_data_combined)

ggsurvplot(km_pfs, data = pfs_data_combined, risk.table = TRUE, pval = TRUE,
           xlab = "Time (days)", ylab = "Progression-free survival probability",
           title = "PFS: ADC vs. No ADC", legend.title = "Arm")

km_os <- survfit(Surv(os_time, os_event) ~ arm, data = pfs_data_combined)

ggsurvplot(km_os, data = pfs_data_combined, risk.table = TRUE, pval = TRUE,
           xlab = "Time (days)", ylab = "Overall survival probability",
           title = "OS: ADC vs. No ADC", legend.title = "Arm")

pfs_data_combined %>%
  group_by(arm) %>%
  summarise(median_pfs = median(pfs_time), median_os = median(os_time),
            pfs_event_rate = mean(pfs_event), os_event_rate = mean(os_event))
pfs_data_combined %>% filter(arm == "ADC") %>% summarise(never_progressed = mean(progression_event == 0))
```

## 10. Conclusion and limitations

This notebook connects the two previously separate strands of the project: the
multi-lesion RECIST framework (notebook 3) and the ADC PK/PD mechanism (notebook 2),
extended here to repeated q3w dosing via the pharmacokinetic superposition principle.
The calibrated model shows a clear, non-trivial treatment effect on both PFS
(p < 0.0001) and OS (p = 0.021), with a more pronounced effect on PFS than OS — a
pattern consistent with real oncology trials, where progression is often more
sensitive to treatment effect than overall survival, which is influenced by
subsequent lines of therapy and competing causes of death not modeled here.

**Limitations, stated explicitly:**

- `Emax` was calibrated empirically (by iterative adjustment) rather than derived from
  a mechanistic or literature-based value, specifically to produce a realistic range of
  treatment response (neither near-0% nor near-100% non-progression) — a pragmatic
  choice made transparent here rather than presented as a validated estimate. An
  initial single-dose version of this model (as in notebook 2) produced a treatment
  effect that faded well before the end of follow-up, diluting any detectable impact on
  PFS/OS over a long observation window — motivating the repeated-dosing structure used
  here.
- Dosing continues at fixed q3w intervals throughout the full follow-up window,
  regardless of whether RECIST progression has already been detected — in reality,
  treatment is typically discontinued or switched at progression. Implementing this
  would require restructuring the simulation as a sequential, day-by-day loop per
  patient (checking progression status before administering each new dose), rather
  than the current vectorized approach that computes the full dosing schedule upfront.
  This was left out as a deliberate scope decision: since PFS is already determined at
  the moment of progression itself, the main effect of this simplification would be a
  slight overestimation of drug exposure — and therefore treatment effect — on OS
  specifically, for the subset of patients who progress before the end of follow-up.
- As in notebook 3, new lesion appearance is not modeled; only growth/shrinkage of
  existing target lesions is considered.
- As in notebook 2, all patients follow the same fixed q3w schedule; real-world
  dose modifications, delays, and discontinuations due to toxicity are not modeled.
